---
title: "Study of the health and economic impacts of severe weather events"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Synopsis

This sudy...

## Data Processing

Load the NOAA Storm data into a data frame
```{r cache=TRUE}
data <- read.csv("repdata_data_StormData.csv.bz2", header=TRUE, quote="\"", sep=",")
```

We will look at the FATALITIES and INJURIES fields to estimate the impact of the weather events of population health we will look at the fields.

To estimate the economic impact of these events, we will look at the damage to crop and properties. The total damage to crop and property can be calculated as follows: 
DAMAGE = PROPDMG * 10 ^ PROPDMGEXP + CROPDMG * 10 ^ CROPDMGEXP

The PROPDMGEXP and CROPDMGEXP however, need to be converted into a numerical value. This is the purpose of the exp_code_to_num function.


```{r}
exp_code_to_num <- function(code) {
  num_exp <- as.integer(0)
  num_exp <- ifelse(code %in% c("h", "H"), 2, num_exp)
  num_exp <- ifelse(code %in% c("+", "k", "K"), 3, num_exp)
  num_exp <- ifelse(code %in% c("M", "m"), 6, num_exp)
  num_exp <- ifelse(code %in% c("B", "b"), 9, num_exp)
  num_exp <- ifelse(code %in% c("?", "-"), 1, num_exp)
  num_exp <- ifelse(code %in% c(""), 0, num_exp)
  num_exp
}

data$PROPDMGEXPNUM <- exp_code_to_num(data$PROPDMGEXP) 
data$CROPDMGEXPNUM <- exp_code_to_num(data$CROPDMGEXP) 

data$DAMAGE <- data$PROPDMG * 10 ^ data$PROPDMGEXPNUM + data$CROPDMG * 10 ^ data$CROPDMGEXPNUM

event_types <- aggregate(cbind(FATALITIES, INJURIES, DAMAGE) ~ EVTYPE, data = data, sum)
```


## Results

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}
event_types$EVTYPE <- reorder(event_types$EVTYPE, -event_types$FATALITIES)
top_events <- order(event_types$FATALITIES, decreasing = TRUE)[1:20]

ggplot(event_types[top_events,], aes(x = EVTYPE, y = FATALITIES)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  xlab("Event Types") +
  ylab("Fatalities") +
  ggtitle("Number of deaths for various type of weather events in the United States")
  
```



